---
title: "Initial analysis and checking"
output: html_document
date: "2025-04-01"
author: "Ishika Verma"
---
# Importing the required libraries
```{r}
library("DESeq2")
library("apeglm")
library("pheatmap")
library("RColorBrewer")
library("pheatmap")
library("vsn")
library(org.Mm.eg.db)
library(ggplot2)
library(EnhancedVolcano)
library(ggVennDiagram)
library(grid)
library(devtools)
library(enrichR)
```


## Reading the files for all the data

```{r}
#HSC files
hsc_1 <- read.table("data/1_HSC/rna/ENCFF064MKY.tsv", header = TRUE, sep = "\t")
hsc_2 <- read.table("data/1_HSC/rna/ENCFF247FEJ.tsv", header = TRUE, sep = "\t")

#CMP files
cmp_1 <- read.table("data/2_CMP/rna/ENCFF623OLU.tsv", header = TRUE, sep = "\t")
cmp_2 <- read.table("data/2_CMP/rna/ENCFF691MHW.tsv", header = TRUE, sep = "\t")

#CFUE files
cfue_1 <- read.table("data/3_CFUE/rna/ENCFF655LMK.tsv", header = TRUE, sep = "\t") 
cfue_2 <- read.table("data/3_CFUE/rna/ENCFF667IDY.tsv", header = TRUE, sep = "\t")  

#ERY files
ery_1 <- read.table("data/4_ERY/rna/ENCFF342WUL.tsv", header = TRUE, sep = "\t") 
ery_2 <- read.table("data/4_ERY/rna/ENCFF858JHF.tsv", header = TRUE, sep = "\t") 

```

# Initial Analysis
This analysis is done to check whether there are same number of genes in all the samples. 

## Extract expected_count along with genes from all the samples
```{r}
#HSC files
hsc_1_counts <- hsc_1$expected_count
hsc_2_counts <- hsc_2$expected_count

#CMP files
cmp_1_counts <- cmp_1$expected_count
cmp_2_counts <- cmp_2$expected_count

#CFUE files
cfue_1_counts <- cfue_1$expected_count
cfue_2_counts <- cfue_2$expected_count  

#ERY files
ery_1_counts <- ery_1$expected_count
ery_2_counts <- ery_2$expected_count
```

## Extract gene_id from all the samples
```{r}
#HSC files
hsc_1_gene_id <- hsc_1$gene_id
hsc_2_gene_id <- hsc_2$gene_id

#CMP files
cmp_1_gene_id <- cmp_1$gene_id
cmp_2_gene_id <- cmp_2$gene_id

#CFUE files
cfue_1_gene_id <- cfue_1$gene_id
cfue_2_gene_id <- cfue_2$gene_id

#ERY files
ery_1_gene_id <- ery_1$gene_id
ery_2_gene_id <- ery_2$gene_id
```

## Ensuring that the length expectd_counts of the samples is the same
```{r}
#HSC files
length(hsc_1$expected_count)
length(hsc_2$expected_count)

#CMP files
length(cmp_1$expected_count)
length(cmp_2$expected_count)

#CFUE files
length(cfue_1$expected_count)
length(cfue_2$expected_count)

#ERY files
length(ery_1$expected_count)
length(ery_2$expected_count)

```

#Ensuring that there are same number of genes in each sample
```{r}
#HSC files
length(hsc_1$gene_id)
length(hsc_2$gene_id)

#CMP files
length(cmp_1$gene_id)
length(cmp_2$gene_id)

#CFUE files
length(cfue_1$gene_id)
length(cfue_2$gene_id)

#ERY files
length(ery_1$gene_id)
length(ery_2$gene_id)
```










