---
title: "RNA seq analysis of CMP vs ERY"
output: html_document
date: "2025-04-20"
author: "Ishika Verma"
---
# Importing the required libraries
```{r}
library("DESeq2")
library("apeglm")
library("pheatmap")
library("RColorBrewer")
library("pheatmap")
library("vsn")
library(org.Mm.eg.db)
library(ggplot2)
library(EnhancedVolcano)
library(ggVennDiagram)
library(grid)
library(devtools)
library(enrichR)
```

## Reading the files for all the data

```{r}
#CMP files
cmp_1 <- read.table("data/2_CMP/rna/ENCFF623OLU.tsv", header = TRUE, sep = "\t")
cmp_2 <- read.table("data/2_CMP/rna/ENCFF691MHW.tsv", header = TRUE, sep = "\t")
cmp_1_counts <- cmp_1$expected_count
cmp_2_counts <- cmp_2$expected_count

#ERY files
ery_1 <- read.table("data/4_ERY/rna/ENCFF342WUL.tsv", header = TRUE, sep = "\t") 
ery_2 <- read.table("data/4_ERY/rna/ENCFF858JHF.tsv", header = TRUE, sep = "\t") 
ery_1_counts <- ery_1$expected_count
ery_2_counts <- ery_2$expected_count

```

# Pair 3 - CMP & ERY

# Checking whether the gene ids are identical before making a count matrix with them

```{r}
identical(cmp_1$gene_id, cmp_2$gene_id)
identical(ery_1$gene_id, ery_2$gene_id)
identical(cmp_1$gene_id, ery_1$gene_id)
identical(cmp_2$gene_id, ery_2$gene_id)
```
Since all the outputs are true, it means that the gene_ids are same in all the samples of cmp and ery.

#Making the DESeqDataset (dds)
```{r}
#countData
cmp_ery_count_data <- data.frame(
  cmp_1 = cmp_1_counts,
  cmp_2 = cmp_2_counts,
  ery_1 = ery_1_counts,
  ery_2 = ery_2_counts,
  row.names = cmp_1$gene_id
)

#colData
samples <- c("cmp_1", "cmp_2", "ery_1", "ery_2")
broad_category_samples <-rep(c("cmp","ery"),each=2)
cmp_ery_coldata <- data.frame(row.names = samples, cell_stage = broad_category_samples)

#dds
dds_cmp_ery <- DESeqDataSetFromMatrix(countData = round(cmp_ery_count_data),
                             colData = cmp_ery_coldata,
                             design = ~ cell_stage )
```
```{r}
#Saving the count matrix in a csv file
write.csv(cmp_ery_count_data, file = "24_cmp_ery_count_data.csv", quote = FALSE)
```

#Find the smallest group size and performing pre-filtering
This step ensures that the genes with low read count are removed from the data prior to 
```{r}
table(dds_cmp_ery$cell_stage)
smallestGroupSize <- 2
keep <- rowSums(counts(dds_cmp_ery) >= 10) >= smallestGroupSize
dds_cmp_ery <- dds_cmp_ery[keep,]
```


## Performing differential gene expression analysis
```{r}
dds_cmp_ery <- DESeq(dds_cmp_ery)
res_cmp_ery <- results(dds_cmp_ery)
summary(res_cmp_ery)
```

## Log fold change shrinkage
```{r}
resultsNames(dds_cmp_ery)
resLFC_cmp_ery <- lfcShrink(dds_cmp_ery, coef="cell_stage_ery_vs_cmp", type="apeglm")
resLFC_cmp_ery
```

## Ordering the top genes by p-values
```{r}
resOrdered <- resLFC_cmp_ery[order(resLFC_cmp_ery$pvalue),]
levels(dds_cmp_ery$cell_stage)
```

## Checking the number of the significant genes (with adjusted p-values less than 0.05)
```{r}
sum(resLFC_cmp_ery$padj < 0.05, na.rm=TRUE)
```

#Mapping ENSEMBL Ids to gene symbols
```{r}
clean_ids <- sub("\\..*", "", rownames(resLFC_cmp_ery))
clean_ids_dds <- sub("\\..*", "", rownames(res_cmp_ery))

gene_symbols <- mapIds(org.Mm.eg.db,
                       keys = clean_ids,
                       column = c('SYMBOL'),
                      keytype = "ENSEMBL")
gene_symbols_dds <- mapIds(org.Mm.eg.db,
                       keys = clean_ids_dds,
                       column = c('SYMBOL'),
                       keytype = "ENSEMBL")



gene_symbols <- gene_symbols[match(clean_ids, names(gene_symbols))]
gene_symbols_dds <- gene_symbols_dds[match(clean_ids_dds, names(gene_symbols_dds))]

resLFC_cmp_ery$symbol <- gene_symbols
res_cmp_ery$symbol <- gene_symbols_dds
```
## Volcano Plots
```{r}
png("CMP_vs_ERY_volcano_deseq2.png", width = 2000, height = 1600, res = 300)
EnhancedVolcano(res_cmp_ery,
                lab = res_cmp_ery$symbol, #not shrunken
                x = 'log2FoldChange',
                y = 'padj',
                xlim = c(-5, 5),
                pCutoff = 0.05,
                FCcutoff = 2,
                title = 'Volcano plot: CMP vs ERY',
                subtitle = 'DESeq2',
                caption = 'Thresholds: padj < 0.05, |log2FC| > 1',
                pointSize = 2.0,
                labSize = 3.0,
                colAlpha = 0.8,
                legendPosition = 'right',
                legendLabSize = 12,
                legendIconSize = 4.0, 
                ylab = "-Log10(padj)",
  legendLabels = c("NS", "Log2 FC", "padj", "padj and log2 FC"))
dev.off()
```


# Saving upregulated and downregulated genes for functional enrichment analysis 
```{r}
top_genes_logfc_1 <- subset(res_cmp_ery, padj < 0.05 & abs(log2FoldChange) >= 1)

up_genes <- subset(top_genes_logfc_1, log2FoldChange >= 1)
down_genes <- subset(top_genes_logfc_1, log2FoldChange <= -1)

up_ensembl_ids <- sub("\\..*", "", rownames(up_genes))
down_ensembl_ids <- sub("\\..*", "", rownames(down_genes))

write.table(up_ensembl_ids,
            file = "up_genes_logfc1_cmp_vs_ery_deseq2_ensembl_ids.txt",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)

write.table(down_ensembl_ids,
            file = "down_genes_logfc1_cmp_vs_ery_deseq2_ensembl_ids.txt",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)
```

##MA Plot
```{r}
plotMA(resLFC_cmp_ery, alpha = 0.05)
title("MA Plot: CMP vs ERY")
```



## Clustering analysis
### Transformation
```{r}
vsd_cmp_ery <- vst(dds_cmp_ery, blind=FALSE)
```

## Heatmap of count matrix
```{r}
library("pheatmap")
select <- order(rowMeans(counts(dds_cmp_ery,normalized=TRUE)),
                decreasing=TRUE)[1:50]
df <- as.data.frame(colData(dds_cmp_ery)[, "cell_stage", drop = FALSE])
pheatmap(assay(vsd_cmp_ery)[select,], cluster_rows=FALSE, show_rownames=FALSE,
         cluster_cols=FALSE, annotation_col=df,  main = "Top 50 Most Expressed Genes in CMP vs ERY")



```
### Sample clustering
### Calculate sample-to-sample distances
```{r}
sampleDists_cmp_ery <- dist(t(assay(vsd_cmp_ery)))

sampleDistMatrix_cmp_ery <- as.matrix(sampleDists_cmp_ery)
rownames(sampleDistMatrix_cmp_ery) <- paste(vsd_cmp_ery$cell_stage, vsd_cmp_ery$type, sep="-")
colnames(sampleDistMatrix_cmp_ery) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Greys")) )(255)
pheatmap(sampleDistMatrix_cmp_ery,
         clustering_distance_rows=sampleDists_cmp_ery,
         clustering_distance_cols=sampleDists_cmp_ery,
         col=colors,
         main = "Sample-to-Sample Distance Matrix (CMP vs ERY)")
```


